---
title: 'Split multiple response entry column(s) in R'
author: "Steve Cygu ([cygubicko@gmail.com](mailto:cygubicko@gmail.com))"
date: "2019 Feb 21 (Thu)"
output:
  html_document:
   code_folding: show
   number_sections: yes
   toc: yes
   toc_depth: 5
   toc_float: yes
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE
	, warning = FALSE
	, message = FALSE
	, fig.width = 10
	, fig.height = 10
	, results = "asis")
options(width = 12)
```

# Setup

We begin by loading the required packages. If you don't have the packages installed, use `install.packages("package_name")`.

```{r, echo = TRUE}
library(DT)
library(dplyr)
library(tidyr)
```


# Problem

In many data collection software, multiple response questions' selections (respondent can choose more than one option) are usually stored in a single cell separated with some characters. This problem is mostly common in _ODK_ and other _ODK_ based tools or even Google Forms. In addition, a more general problem could be creating columns from a single column of a dataframe based on some pattern.

To demonstrate how we go about this, we use [social media survey data](https://github.com/CYGUBICKO/miscellaneous/tree/master/datasets) in which reposndents were asked which social media they use. Let us take a look at the variables.

```{r}
desc <- read.csv("../datasets/multi_resp_desc.csv")
datatable(desc, rownames = FALSE)
```

In this example, the respondents could have selected more than one social media platforms (see **smedia\_used**). The selected oprions are seperated using "," and our aim is to create additional variables to hold each of the selected responses. Let us take a look at the data

```{r, echo = TRUE}
smedia_df <- read.csv("../datasets/multi_response.csv")
datatable(head(smedia_df), rownames = FALSE)
```

Some of the ways to solve thus problem may include:

* Split columns in excel

* Use R functions

In this post, we will show a step by step guide on how to write a R function which will be able to do the job.


# R way

As mentioned above, we are going to demonstrate this using **smedia\_used** variable. We would therefore ask ourselves how much information do we have and best can use such?

|What we see | What R will understand | Comment |
|------- | ------- | -------------|
|","     | pattern | This can be any pattern ("-", "space",...) dependeing on the plattform/problem.| 
|Individual entries in **smedia\_used** | Number of new variables to create | Number of new variables to be created equals max(number of entries).|
----------------------------------------------------------------------
 

First we need to understand how pattern matching and regular expressions work in R. You can find some help online but I would recommend [Regular Expressions in R](https://rstudio-pubs-static.s3.amazonaws.com/74603_76cd14d5983f47408fdf0b323550b846.html) or [Pattern Matching and Replacement](http://astrostatistics.psu.edu/su07/R/html/base/html/grep.html) as a starting point.

We need to count the number of selections in **smedia\_used** based on the pattern ",".

```{r, echo = TRUE}
countFunc <- function(var, pattern){
	nselect <- sapply(regmatches(var, gregexpr(pattern, var)), length) + 1
	return(nselect)
}
```

* `var` - input variable (could be multiple response)
* `pattern` - what separates the entries. Could be ",", "\_" ....
* `gregexpr()` and `regmatches()` - Searches for the pattern and returns the position if TRUE.
* `sapply` - tells R to 'repeatedly' search for the pattern and do something - _length_.
* `length` - what _sapply_ does. Counts the number of patterns found.
* `+1` - counting only happens if there is more than one entry. This accounts for first selection.

Let us test the `countFunc`.
```{r, echo = TRUE}
test_vec <- c("first, second, third", "fourth", "second, fifth")
pattern <- ","
countFunc(test_vec, pattern)
```

We've correctly counted the entries in _test_$\_$_vec_. Now let us apply the function, `countFunc()`, to our dataset.

```{r, echo = TRUE}
x <- pull(smedia_df, "smedia_used")
pattern <- ","
nselected <- countFunc(x, pattern) # Count the number of entries

# Add this column to the dataset
smedia_df2 <- (smedia_df
   %>% mutate(nselected = nselected)
)
datatable(smedia_df2, rownames = FALSE)
```

The last column in the output above correctly counts the number of selections in **smedia**\_**df**. We now modify `countFunc()` to perform all the tasks above and add a few lines to split the multiple responses into new columns.

```{r, echo = TRUE}
countFunc <- function(df, var, pattern, remove = FALSE, drop_ns = FALSE){
	nselected <- paste0(var, "_nselected")
	df <- (df
		%>% rename(temp_multi = var)
		%>% mutate(temp_nselected = sapply(
				regmatches(temp_multi, gregexpr(pattern, temp_multi))
				, length
			) + 1
		)
	)
	# Max number of new variables to create
	maxselected <- max(pull(df, temp_nselected))

	# Create the new variable
	df <- (df 
		%>% separate_("temp_multi"
			, c(paste0(rep(var, maxselected), "_", 1:maxselected))
			, sep = pattern
			, remove = remove
			, convert = TRUE
		)
		%>% rename_(.dots = setNames(c("temp_multi", "temp_nselected"), c(var, nselected)))
	)
	if(drop_ns){
		df <- select(df, -c(grep("_nselected$", colnames(df), value = TRUE)))
	}
	return(df)
}
```

This function might look complicated but not at all...! 

* `var` and `pattern` - same as above.
* `remove = TRUE` - drops the parent multiple response variable.
* `drop_ns` - whether to create a variable for the number of selections per case. Naming is of the form `varname_nselected`.
* `mutate()` - counts the number of selected reponses and then create a temporary variable, `temp_nselected`.
* `separate()` - based on the pattern and the maximum number of selections made, this line creates new columns immediately after the multiple variable. The created variables keeps same variable name with "\_" number.

Let us take a look at the output

```{r, echo = TRUE}
smedia_df3 <- (smedia_df
	%>% countFunc(var = "smedia_used", pattern = ",", remove = FALSE, drop_ns = FALSE)
	%>% datatable(., rownames = FALSE, options = list(scrollX = TRUE))
)
smedia_df3
```

* Try `remove = TRUE` and/or `drop_ns = TRUE`.

Technically, we've achieved our aim; counted the selections and managed to create individual variables. However, we can put everything together and make our function more robust.

## Everything together

Suppose we want to split several variable? Of course we can simply do a `for loop` but it would be much better if we implemented that in `countFunc` so that we don't have think about _how to_ and maybe give it a different name like `multiSeparate`.

```{r, echo = TRUE}
multiSeparate <- function(df, vars, pattern, remove = FALSE, drop_ns = FALSE){
	for (var in vars){
		nselected <- paste0(var, "_nselected")
		df <- (df
			%>% rename(temp_multi = var)
			%>% mutate(temp_nselected = sapply(
					regmatches(temp_multi, gregexpr(pattern, temp_multi))
					, length
				) + 1
			)
		)
		# Max number of new variables to create
		maxselected <- max(pull(df, temp_nselected))

		# Create the new variable
		df <- (df 
			%>% separate_("temp_multi"
				, c(paste0(rep(var, maxselected), "_", 1:maxselected))
				, sep = pattern
				, remove = remove
				, convert = TRUE
			)
			%>% rename_(.dots = setNames(c("temp_multi", "temp_nselected"), c(var, nselected)))
		)
		if(drop_ns){
			df <- select(df, -c(grep("_nselected$", colnames(df), value = TRUE)))
		}
	}
	return(df)
}
```

* We only added 2 lines and changed one input. Your guess is as good as mine.

Test our function we create $3$ additional dummy multiple variables with different separation patterns.

```{r, echo = TRUE}
smedia_df <- (smedia_df
 %>% mutate(smedia_used_dummy1 = smedia_used
 		, smedia_used_dummy2 = gsub(",", "-", smedia_used)
 		, smedia_used_dummy3 = gsub(",", " ", smedia_used)
 )
)

multi_vars <- grep("^smedia_used", colnames(smedia_df), value = TRUE)
print("All multiple respnse variables")
multi_vars
```

```{r, echo = TRUE}
smedia_df <- (smedia_df
	%>% multiSeparate(vars = multi_vars, pattern = ",| |-", remove = FALSE, drop_ns = FALSE)
	%>% datatable(., rownames = FALSE, options = list(scrollX = TRUE))
)
smedia_df
```

You can download the function from [here](https://github.com/CYGUBICKO/miscellaneous/tree/master/funs) or markdown file from [here](https://github.com/CYGUBICKO/miscellaneous/tree/master/data_mgt).
