---
title: 'Split multiple response entry column(s)'
author: "Steve Cygu ([cygubicko@gmail.com](mailto:cygubicko@gmail.com))"
date: "2019 Feb 21 (Thu)"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 5
    toc_float: yes
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = FALSE
	, warning = FALSE
	, message = FALSE
	, fig.width = 10
	, fig.height = 10
	, results = "asis")
options(width = 12)
```

# Setup

We begin by loading the required packages. If you don't have the packages installed, use `install.packages("package_name")`.

```{r, echo = TRUE}
library(DT)
library(dplyr)
```


# Problem

In many data collection software, multiple response questions' selections (respondent can choose more than one option) are usually stored in a single cell separated with some characters. This problem is mostly common in _ODK_ and other _ODK_ based tools or even Google Forms. In addition, a more general problem could be creating columns from a single column of a dataframe based on some pattern.

To demonstrate how we go about this, we use [social media survey data](https://github.com/CYGUBICKO/miscellaneous/tree/master/datasets) in which reposndents were asked which social media they use. Let us take a look at the variables.

```{r}
desc <- read.csv("../datasets/multi_resp_desc.csv")
datatable(desc, rownames = FALSE)
```

In this example, the respondents could have selected more than one social media platforms (see **smedia\_used**). The selected oprions are seperated using "," and our aim is to create additional variables to hold each of the selected responses. Let us take a look at the data

```{r, echo = TRUE}
smedia_df <- read.csv("../datasets/multi_response.csv")
datatable(head(smedia_df), rownames = FALSE)
```

Some of the ways to solve thus problem may include:

* Split columns in excel

* Use R functions

In this post, we will show a step by step guide on how to write a R function which will be able to do the job.


# R way

As mentioned above, we are going to demonstrate this using **smedia\_used** variable. We would therefore ask ourselves how much information do we have:

|What we see | What R will understand | Comment |
|------- | ------- | -------------|
|","     | pattern | This can be any pattern ("-", "space",...) dependeing on the plattform/problem.| 
|Individual entries in **smedia\_used** | Number of new variables to create | Number of new variables to be created equals max(number of entries).|
