---
title: 'Summariz(s)e all the variables in a dataframe'
author: "Steve Cygu ([cygubicko@gmail.com](mailto:cygubicko@gmail.com))"
date: "2019 Feb 28 (Thu)"
output:
  html_document:
   code_folding: show
   number_sections: yes
   toc: yes
   toc_depth: 5
   toc_float: yes
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE
	, warning = FALSE
	, message = FALSE
	, fig.width = 10
	, fig.height = 10
	, results = "asis")
options(width = 12)
```
# Setup

```{r}
library(xtable)
library(DT)
library(dplyr)
```

# Problem

There are many functions or packages which summarizes variables in a data frame. However, there seem not to offer so much flexibility on how much one can manipute or export the output, for instance, including outputs in .tex files. 

In this post, we implement a simple function which entirely depends on base R to generate a more flexible variable(s) summary object.

```{r}
summarizeDf <- function(df, output = c("simple", "tex"), digits = 1){
	if (!missing(output) & sum(!output %in% c("simple", "tex")) > 0){
      stop("output can only be 'simple' or 'tex'")
   }
   vars <- colnames(df)
   df_summary <- data.frame(Variable = rep(NA, length(vars))
      , Type = rep(NA, length(vars))
      , Summary = rep(NA, length(vars))
   )
   for (i in 1:length(vars)){
      vals <- df[, vars[[i]]]
      if (class(vals) == "numeric" | class(vals) == "integer"){
         df_summary[["Type"]][[i]] <- "numeric"
         df_summary[["Variable"]][[i]] <- vars[[i]]
         df_summary[["Summary"]][[i]] <- paste0("["
            , round(min(vals), digits), ", "
            , round(max(vals), digits), "]; "
            , round(mean(vals), digits), " ("
            , round(sd(vals), digits), ")"
         )
      } else{
         df_summary[["Type"]][[i]] <- "categorical"
         df_summary[["Variable"]][[i]] <- vars[[i]]
         perc <- sort(round(prop.table(table(vals))*100, digits)
            , decreasing = TRUE
         )
         if (missing(output) | sum(output %in% "simple") > 0){
            perc <- paste0(names(perc), " (", perc, "%)")
            df_summary[["Summary"]][[i]] <- paste0(perc
               , collapse = ";\n "
            )
         } else{
            perc <- paste0(names(perc), " (", perc, "\\%)")
            df_summary[["Summary"]][[i]] <- paste0(perc
               , collapse = "; \\\\ & & "
            )
         }
      }
   }
   return(df_summary)
}

```

`summarizeDf` summariz(s)es dataframe. Computes ([min, max]; mean (sd)) for numerical or integer variables and frequency distribution (percent) for categorical variables.

Inputs:

> * `df` - Input dataframe
> * `output` - Specifies the output structure. `output = "simple"` returns R-output-like output. `output = "tex"` returns xtable ready format.
> * `digits` - Number of digits to return.

Details:

> For categorical variables with several categories, `output = "tex"` is preferrable. Add `sanitize.text.function = function(x){x}` to xtable print function for .tex.

Value:

> It returns an object of class `data.frame`. Computes ([min, max]; mean (sd)) for numerical or integer variables and frequency distribution (percent) for categorical variable

# Example(s)

To demonstrate this, we use [social media survey data](https://github.com/CYGUBICKO/miscellaneous/tree/master/datasets) described in this [post](https://cygubicko.github.io/miscellaneous/separate_entries.html).


```{r}
smedia_df <- read.csv("../datasets/multi_response.csv")
datatable(smedia_df, rownames = FALSE)
```

Output normal R-like summary. See value above.

```{r}
smedia_df1 <- select(smedia_df, -c("doi"))
smedia_summary <-(smedia_df1
  	%>% summarizeDf(.)
)
datatable(smedia_summary, rownames = FALSE)
```

Generate simple latex-like table.

```{r}
smedia_summary <-(smedia_df1
  	%>% summarizeDf(., output = "simple")
	%>% xtable(., caption = "Simple data summary")
)
summary.tex <- print(smedia_summary, sanitize.text.function = function(x){x}
	, type = "html"
	, scalebox = 0.5
  	, include.rownames = FALSE
  	, caption.placement = "top"
)
```

You can also generate .tex file to include in .tex files by changing `type = "latex"` and also adding `file = "filename.tex"`. See `print.xtable`.

You can download the function from [here](https://github.com/CYGUBICKO/miscellaneous/tree/master/funs) or markdown file from [here](https://github.com/CYGUBICKO/miscellaneous/tree/master/data_mgt).


