---
title: 'Some R Hacks'
author: "Steve Cygu ([cygubicko@gmail.com](mailto:cygubicko@gmail.com))"
date: "2019 Feb 28 (Thu)"
output:
  html_document:
   code_folding: show
   number_sections: yes
   toc: yes
   toc_depth: 5
   toc_float: yes
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE
	, warning = FALSE
	, message = FALSE
	, fig.width = 10
	, fig.height = 10
	, results = "asis")
options(width = 12)
```
# Setup

```{r}
pkgs <- c("xtable", "DT", "dplyr", "tibble", "data.table"     # Reading and saving data
)
if (!"pacman" %in% installed.packages()[,1]){
  install.packages("pacman")
}
pacman::p_load(pkgs, install = T, character.only = T)

```

Focus will be on... but we can also think through some other problems!!!

* Summariz(s) variables in a data frame
* Split multiple response column
* Organise/Re-order variables

Read data
```{r}
file <- list.files()
patterns <- "multi_resp_desc.csv"

if (length(file)>0 & sum(grepl(patterns, file)) > 0){
  df <- read.csv(patterns)
} else {
  df <- read.csv("https://github.com/CYGUBICKO/miscellaneous/raw/master/datasets/multi_response.csv")
}

```

# Check missing values

```{r}

## Missing values
var1 <- is.na(df$doi)
# vars <- names(df)
# na_list <- list()
# for(var in vars){
#   na_list[[var]] <- is.na(df[, var])
# }
# data.frame(na_list)

miss_df <- (sapply(df, function(x){sum(is.na(x))})
  %>% data.frame()
  %>% rownames_to_column("variables")
  %>% setnames(names(.), c("variables", "No_miss"))
  %>% datatable()
)

miss_df

```

```{r}
missFunc <- function(df_frame){
  miss_df <- (sapply(df_frame, function(x){sum(is.na(x))})
    %>% data.frame()
    %>% rownames_to_column("variables")
    %>% setnames(c("."), c("No_miss"))
  )
  return(miss_df)
}
miss_summary <- missFunc(df)
datatable(miss_summary)
```

```{r}

## Create a new dataset df2 with a missing case to test missFunc
df2 <- (df
  %>% mutate(gender2 = ifelse(gender=="Female", NA, gender))
)

# Check the missing in gender2
missFunc(df2)

## Age
df2 <- (df
  %>% mutate(age2 = ifelse(age > 25, NA, age))
)

missFunc(df2)

```

## Summarize dataframes

# Problem

There are many functions or packages which summarizes variables in a data frame. However, there seem not to offer so much flexibility on how much one can manipute or export the output, for instance, including outputs in .tex files.

```{r}
#### ---- DF summary ----
vars <- names(df2)
df_summary <- data.frame(Variables = rep(NA, length(vars))
  , Type = rep(NA, length(vars))
  , Summary = rep(NA, length(vars))
)
df_frame <- df2
digs <- 2
for (i in 1:length(vars)){
  df_summary[["Variables"]][[i]] <- vars[[i]]
  vals <- df_frame[, vars[[i]]]
  if (class(vals) == "factor" | class(vals)== "character"){
    df_summary[["Type"]][[i]] <- "Categorical"
  } else {
    df_summary[["Type"]][[i]] <- "numeric"
    df_summary[["Summary"]][i] <- paste0("[" 
      , round(min(vals, na.rm = TRUE), digits = digs), ","
      , round(max(vals, na.rm = TRUE), digits = digs), ","
      , round(mean(vals, na.rm = TRUE), digits = digs)
    , "]"
    , "("
    ,  round(sd(vals, na.rm = TRUE), digits = digs)
    , ")"
    )
  }
}
df_summary
```

In this post, we implement a simple function which entirely depends on base R to generate a more flexible variable(s) summary object.

```{r}
summarizeDf <- function(df, output = c("simple", "tex"), digits = 1){
	if (!missing(output) & sum(!output %in% c("simple", "tex")) > 0){
      stop("output can only be 'simple' or 'tex'")
   }
   vars <- colnames(df)
   df_summary <- data.frame(Variable = rep(NA, length(vars))
      , Type = rep(NA, length(vars))
      , Summary = rep(NA, length(vars))
   )
   for (i in 1:length(vars)){
      vals <- df[, vars[[i]]]
      if (class(vals) == "numeric" | class(vals) == "integer"){
         df_summary[["Type"]][[i]] <- "numeric"
         df_summary[["Variable"]][[i]] <- vars[[i]]
         df_summary[["Summary"]][[i]] <- paste0("["
            , round(min(vals), digits), ", "
            , round(max(vals), digits), "]; "
            , round(mean(vals), digits), " ("
            , round(sd(vals), digits), ")"
         )
      } else{
         df_summary[["Type"]][[i]] <- "categorical"
         df_summary[["Variable"]][[i]] <- vars[[i]]
         perc <- sort(round(prop.table(table(vals))*100, digits)
            , decreasing = TRUE
         )
         if (missing(output) | sum(output %in% "simple") > 0){
            perc <- paste0(names(perc), " (", perc, "%)")
            df_summary[["Summary"]][[i]] <- paste0(perc
               , collapse = ";\n "
            )
         } else{
            perc <- paste0(names(perc), " (", perc, "\\%)")
            df_summary[["Summary"]][[i]] <- paste0(perc
               , collapse = "; \\\\ & & "
            )
         }
      }
   }
   return(df_summary)
}

df_summary <- (df2
	%>% select(-doi)
	%>% summarizeDf("tex", 2)
	%>% datatable()
)
df_summary
```

This is a collection of scripts to solve some particular problem(s).Some could be thinking process through ideas or otherwise!!!

* [Summariz(s) variables in a data frame](https://cygubicko.github.io/miscellaneous/summarize_variables.html)
* [Split multiple response column](https://cygubicko.github.io/miscellaneous/separate_entries.html)
* [Organise/Re-order variables](https://cygubicko.github.io/miscellaneous/organise_variables.html)

